On Devon Allen's Disqualification at the 2022 Track and Field World Championship

### 2023-04-27 Manuscript/manuscript.tex
Went over some of Prof Schifano's notes

%1)
Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
%After coming in third in the 110 meter hurdles at the United States Track and 
%Field Championships also in Eugene at the end of June, Allen was poised to 
%compete in the 2022 World Track and Field Championships also 
%located at Eugene in August.
%two months later. 
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 in his preliminary heat and 0.101 in his semifinal.  
%Allen, who attended the University of Oregon also in Eugene, 
%(the location of US World Championship 
%\eds{US World? remove World?} and the World Championship) 
%was
%considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
%in early June; only 0.04 seconds away from the world record \citep{wa2022preview}.
%Thus 
His disqualification in the final heat by reacting in 0.099 seconds instead of 
0.1 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification; it can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers 
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting another two years until the next World Track and Field Championship.
\eds{you mention in the final section that the next world championship is in 
2023, so he won't need to wait two years if that is true}
%\eds{final two sentence seems out of place here; consider moving elsewhere?}
%\of{Moved to the end of paragraph 4 of introduction}

%2)
Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit, is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as \citet{johnson2022data}
and \citet{johnson2022was} from  Robert Johnson, who created and runs 
the LetsRun website.  He argued that there was an error
with the timing equipment at the 2022 Track and Field World Championship and
cited graphs, computed simple statistics such as median reaction times, and 
compared the reaction times at the 2022 World Track and Field Championship to 
other competitions
such as the USA Track and Field Championships and other World Championships.
The comparisons lacked formal statistical analysis, however. 
The alleged 99.95\% chance Allen was wrong \eds{can we use different word 
instead of wrong?  at fault?  However, should that be 0.05\% chance at fault?}
came from the fact that the lowest average reaction times for the men's 100 meter,
men's 110 meter hurdle, and women's 100 meter competitions held between 2003 and 
2022 each \eds{?} occurred at the 2022 World 
Championships.  Johnson claimed that the probability of this
happening was $\frac{1}{13} \cdot \frac{1}{13} \cdot \frac{1}{13} = 0.05\%$
\citep{johnson2022was}.
%\eds{compared the reaction times at the 2022 World Track and Field  Championship 
%to what? Also restate here that no formal statistical analyses have been done.}
%\of{Cleaned this up a little but still a little messy.}

@@

%4)
\jy{language needs to be formalized and condensed.}
The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championship to check for any abnormalities, and we approach
this two different ways.  First we model the data, which is
the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with a random effect 
accounting for the venue (or year). \eds{are you using covariates?  Data section
implies that you are, but Results section does not} 
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Second, we compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship.  By
comparing reaction times from the same athlete across multiple competitions we 
are able to determine if the 2022 data differed significantly from other 
competitions.
Our investigation into reaction times is important not only because if its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times; which is the goal of
every sprinter who steps onto the track.
%\eds{I recommend moving the rest of this paragraph to Section 2.2 as I don't
%think it is critical for readers to know all the details here.}
%First we can compare the reaction times of athletes who
%competed at the World Track and Field Championship in both 2019 and 2022, and 
%second we can compare how athletes performed at the 2022 championship relative 
%to their national/qualifiers meet. In order to reduce any chances
%of athletes improving reaction time, we can also look at data from qualifying meets.
%Almost every athlete had to qualify by proving they are elite at the national level, 
%usually by coming in at least fourth. Thus, data was compiled on athletes who competed
%at the 2022 World Track and Field Championships on what these athletes also ran 
%at their country's national championship. If there was any argument that athletes
%have improved their reaction time from 2019 to 2022, it should now be considered
%nearly null, as the national competitions took place between June and July 2022 
%and the World Championships were in August. The equipment used to measure reaction
%time cannot be guaranteed to be the same from national to international
%competition, but the time span between meets has been reduced considerably.

The rest of this paper is organized as follows. Section~\ref{sec:Data} describes 
how data was collected and begins to detail the generalized linear mixed-effect 
(GLMM) model that is further developed in Section~\ref{sec:Methods}. The 
signed-rank test that analyzes reaction times for the same athletes across
different competitions is
also described in Section~\ref{sec:Methods}.  Section~\ref{sec:Results}
 provides the results of the two analyses and ends 
by questioning the reaction time barrier.  Lastly, Section~\ref{sec:Discussion}
discusses the impact and limitations of the paper.
\eds{capitalize first letter when referring to a numbered section, table, figure, 
etc. ``Section 4 provides...'' vs. ``The next section provides...''}

@@

Data was taken from the World Athletics 
%\eds{is there a website you can cite here?}
and covers the men's 110 meter hurdles 
and the women's 100 meter hurdles from 1999 to 2022.  \eds{see earlier comment - 
are you using any of these as covariates?}
The data source can be found in the Appendix.%~\ref{sec:Appendix}. 
Factors such as name, 
country, and identification number were not considered important for the 
purposes of this research besides the times for Devon Allen and the other 
United States athletes mentioned above. \eds{not sure which other US athletes 
you are referring to in the previous sentence} 
The variable ``Stage" refers to whether the observation occurred during a 
%heat (preliminary round)  
preliminary heat denoted by ``H", semifinal heat (of which there are either two
or three each year) denoted by ``S", or final heat denoted by ``F". The variable 
``Gender" has levels ``M" and ``F" for ``male" and ``female" respectively.  The
variable ``Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly). There is no statistical meaning to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary heat of 2022 and Batch
111 corresponds to the 1999 finals). %Included are
%box plots of the reaction times of the men's finals and the men's semifinals and
%finals pooled together.  
For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together and refer to this data as the ``pooled data"
%We will refer to the latter as the ``pooled data"
throughout the paper.  %The heats or preliminary rounds were not studied because
The preliminary heats were not studied because athletes react faster as the 
competition progresses \citep[e.g.,][]{zhang2021correlation}. % Athletes react 
%faster in the finals than the semifinals which are faster than the heats 
%\citep{zhang2021correlation}.  
\eds{if you are really not using ``H'' data at all, it does not appear necessary
to spend time defining earlier.  You could just say directly when you define 
stage that you only consider semifinals and finals and cite the Zhang paper for 
rationale}
If we want to determine if Allen's reaction time
was reasonable, then we thus need to look at the men's final data, with the men's
pooled data being included in order to increase the sample size.  
\eds{previous sentence is not clear; pooled data already includes final data}
In 2022, there
were only five data points due to two disqualifications and one athlete who 
qualified but did not compete.
The data under consideration is best visualized using side-by-side box plots, as
shown in Figure~\ref{fig:Boxplots}.  It is clear that the 2022 box plot for both 
the men's finals and semifinals data is lower (indicating faster reaction times)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
\citet{brosnan2017effects} found reaction times of 0.156 seconds \eds{mean?  median?} for men when looking at data from 1999 to 2014.

\eds{I wonder if this paragraph would be better placed in the introduction?}
Something that is extremely important to note in the data is the inconsistency 
of World Athletics in terms of how they penalize false starts.  From 2007--2009,
the IAAF (the former name for World Athletics, and the governing body for the 
World Track and Field Champions) instituted a rule change that allowed one 
warning false start before a sprinter was disqualified 
\citep{iaaf2009falsestart}.  This 
essentially gave sprinters a second chance and reduced the penalty for a false 
start.  There were 18 male and 7 female false starts at the 2007 World 
Championships, and 18 male and 7 female starts at the 2009 World Championships. 
Starting in 2011, the rule was scrapped, and the old policy which automatically 
disqualifies runners who false start was reinstated. By returning to the harsher 
policy and cracking down on false starts, World Athletics had reduced the number 
of false starts by two thirds \eds{for men, not overall} in 2011 (6 male and 4 female disqualifications 
\citep{iaaf2009falsestart}).
%The effect? At the 2011 
%World Championship there were 6 male disqualifications and 4 female 
%disqualifications \citep{iaaf2009falsestart}.  
%One study that looked at different false start rules that IAAF has
%imposed from 1997 to 2009 and found that the reaction times when there was less 
%leniency improved by a statistically significant amount
%\citep{haugen2013effect}. 
\citet{haugen2013effect} examined the effect of different false start
rules that IAAF has imposed from 1997 to 2009 and found statistically 
significant improvement in reaction times under harsher rules.
\eds{is that right?  it seems counterintuitive to me}
This was desirable for World Athletics as false starts can make already long 
track meets more tedious, both for athletes and viewers watching the 
television broadcast.  
\eds{I am not sure the previous sentence follows from the one preceding it. Why
 does World Athletics find faster reaction times desireable?  Do you mean that
they find fewer false starts desirable?}
Since 2011, IAAF has instituted the one false start rule
\citep{iaaf2009falsestart}. However, we chose not to include women's data in this
paper due to the numerous papers citing difference in reaction times in men and
women's reaction times: \citep{lipps2011implications, babicc2009reaction,
panoutsakopoulos2020gender}  \eds{Two comments about the previous sentence: 1. 
it does not logically follow the previous sentence as there is no mention of 
gender, and 2. if you are not using women's data, don't spend time talking about 
it at the beginning of the data section.  Similar to my comment about 
preliminary heats, you can move this sentence to the beginning of the Data 
section to provide a rationale for why you do not include data for women.}

\subsection{Beyond World Championship}\label{sec:databeyond}
The motivation for looking beyond World Championship data came from an 
initial qualitative \eds{exploratory instead of qualitative?}
analysis examining how United States athletes reacted at the 
2022 United States Track and Field Championships.
\eds{be consistent with how you refer to the US championship:
``United States Track and Field Championships" and ``USA Track and Field 
Championships" are both being used; please pick one and use that throughout}
%This section was motivated from a qualitative analysis that started by looking
%at how United States athletes reacted at the 2022 United States Track and Field
%Championships. 
From June 23-26 2022, the United States held its Track
and Field Championship at Hayward 
Field in Eugene, Oregon to decide who to send to the World Championships 
being held at the same venue in August. %the exact same venue as used 
%for the World Track and Field Championship a month later.  
%Thus, a baseline is able to be established for the four USA athletes who 
Thus, we can establish a baseline for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teammates. \eds{statistically significantly faster? If you used a statistical 
test, it should not be described in this section.  If you didn't use a 
statistical test, you should replace the word `significantly' with something 
else (substantially?)} 
This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  Table~\ref{fig:USAvsWorld} highlights the difference in reaction 
times for Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. 
%``USA" denotes
%the USA Track and Field Championships, ``World" denotes the World Track and Field
%Championships. All numbers listed in the table are the reaction
%time for the respective athletes in seconds.

\begin{table}
\begin{center}
  \caption{Reaction times (in seconds) for four United States Athletes at the 
	2022 USA Track
  and Field World Championships (``USA") and the 2022 World Track and Field 
	Championships (``World"); H denotes preliminary heat, S denotes semifinal, and 
	F denotes final.
  ``N/A" was used for any athlete that did not compete and thus did not register 
  a reaction time. }
  \begin{tabular}{c c c c c c c} 
   \toprule
	 & \multicolumn{3}{c}{USA} & \multicolumn{3}{c}{World} \\
	\cmidrule(lr){2-4}
    \cmidrule(lr){5-7}
   Athlete &  H &  S &  F &  H &  S &  F \\ [0.5ex] 
   \midrule
   Devon Allen & 0.201 & 0.153 & 0.160 & 0.123 & 0.101 & 0.099 \\ 
   Trey Cunningham & 0.186 & 0.185 & 0.182 & 0.115 & 0.120 & 0.109 \\
   Grant Holloway & 0.192 & 0.190 & N/A & 0.147 & 0.128 & 0.124 \\
   Daniel Roberts & 0.181 & 0.200 & 0.183 & 0.179 & N/A & N/A \\ [0.5ex]
   \bottomrule
  \end{tabular}
  \label{fig:USAvsWorld}

@@

@@

To better understand this model, we can identify the gamma model in terms of the
commonly used shape and scale parameters. For a gamma distribution with
shape~$a$ and scale~$b$, we have mean $\mu = ab$ and variance $v = ab^2$. The
variance as a function of the mean $\mu$ in the GLM sense is
$v(\mu) = \mu^2 / a$, which implies that the dispersion parameter is
$\phi = 1 / a$. Therefore, for the gamma mixed-effect model, the conditional
gamma distribution of the reaction time $Y_{ij}$ given the venue effect $z_i$
has shape $a = 1 / \phi$ and scale $b = g^{-1}(\alpha + z_i) \phi$. The marginal
distribution of $Y_{ij}$ is a scale-mixture of gamma distributions, which can be
easily simulated from once the parameters are estimated. A large number of
random numbers generated from the fitted mixture distribution can be used to
approximate the probability of observing a reaction time faster than any given
threshold.  We are specifically interested in the probability of a reaction time
 being less than 0.1 seconds in order to gauge if that is a reasonable 
disqualification barrier.

As for the random effect, it was possible to include a race 
\eds{heat? when I first read this, I thought you meant race as in ethnicity} 
effect, as well.
From our exploration, however, the race effect is very difficult to explain or 
interpret as it means that some heats of races were either
faster or slower and that everyone in the race reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier, however, to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art. Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the race
effect, but are all factors that could potentially impact the venue effect.

@@

%Citation needed in this section
Since 1985 Seiko Holding Corporations has served every World Athletics 
Championship as the official timer \citep{wa2022seiko}.  Since 1985, the 
technology and the ability
to accurately predict measurements: long jump distances, false starts, total 
time, reaction time, etc. have all dramatically improved.  Seiko did not start 
tracking reaction time as an official measurement until 1999 in the men's and 
women's 100 meter \eds{110 for men?} hurdles. 
Seiko regularly updates their equipment so that they provide cutting edge
technology to the World Track and Field Championships, the highest stakes in the 
world of running outside of the Olympics. Seiko's technology for detecting 
reaction times relies on the pressure that athletes exert on the plate when they 
push off.  Their systems measure the time differential
between when the ``gun" goes off to start the race and when the pressure changes 
\citep{wa2022seiko}.  If an athlete has a reaction time under 0.1 seconds, it 
is deemed a 
false start as it is considered that no athlete can react so quickly 
\citep{Seiko-Timing}.  Thus, a reaction time of for example 0.05, suggests that 
the athlete predicted the gun and it was luck that caused their abnormally high 
time.  At the World Championship level, World Athletics and Seiko want to remove 
that element of luck and thus impose the 0.1 second barrier. In 2013, Seiko 
upgraded their timing equipment for sprinting events (100 meter hurdles falls 
under this category) \citep{wa2013backtage} \eds{110 for men?}. Since 1999, the
 highest \eds{highest as in fastest or slowest? Neither seems consistent with 
the boxplots in Figure 1} reaction 
times in the men's 110-meter hurdle were recorded in 2013.  That is not to say
 that the higher 
times were caused by Seiko's equipment, but the two may be related.  It is worth 
noting that prior
to 2022, Seiko again upgraded its technology but for its jump management system.
\eds{I would argue that this paragraph should go in the discussion as there are 
no results presented here}

>>>>

%1)
Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
%After coming in third in the 110 meter hurdles at the United States Track and 
%Field Championships also in Eugene at the end of June, Allen was poised to 
%compete in the 2022 World Track and Field Championships also 
%located at Eugene in August.
%two months later. 
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 in his preliminary heat and 0.101 in his semifinal.  
%Allen, who attended the University of Oregon also in Eugene, 
%(the location of US World Championship 
%\eds{US World? remove World?} and the World Championship) 
%was
%considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
%in early June; only 0.04 seconds away from the world record \citep{wa2022preview}.
%Thus 
His disqualification in the final heat by reacting in 0.099 seconds instead of 
0.1 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification; it can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers 
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championship.
\eds{you mention in the final section that the next world championship is in 
2023, so he won't need to wait two years if that is true}
\of{Resolved}
%\eds{final two sentence seems out of place here; consider moving elsewhere?}
%\of{Moved to the end of paragraph 4 of introduction}

%2)
Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit, is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as \citet{johnson2022data}
and \citet{johnson2022was} from  Robert Johnson, who created and runs 
the LetsRun website.  He argued that there was an error
with the timing equipment at the 2022 Track and Field World Championship and
cited graphs, computed simple statistics such as median reaction times, and 
compared the reaction times at the 2022 World Track and Field Championship to 
other competitions
such as the United States Track and Field Championships and other World Championships.
The comparisons lacked formal statistical analysis, however. 
The alleged 99.95\% chance Allen was wronged \eds{can we use different word 
instead of wrong?  at fault?  However, should that be 0.05\% chance at fault?}
\of{Changed to "wronged"}
came from Johnson claiming there is a $(1/13)^3$ that the fastest reaction
times came during the same championship for the men's 100 meter,
men's 110 meter hurdle, and women's 100 meter for 13 competitions held between 2003 and 
2022 \eds{?} \of{Cleaned up but may need additional work}
\citep{johnson2022was}.
%\eds{compared the reaction times at the 2022 World Track and Field  Championship 
%to what? Also restate here that no formal statistical analyses have been done.}
%\of{Cleaned this up a little but still a little messy.}

@@

%4)
\jy{language needs to be formalized and condensed.}
The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championship to check for any abnormalities, and we approach
this two different ways.  First we model the data, which is
the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with a random effect 
accounting for the venue (or year). \eds{are you using covariates?  Data section
implies that you are, but Results section does not}
\of{Cleaned up confusion about covariates, but no none were used in the paper} 
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Second, we compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship.  By
comparing reaction times from the same athlete across multiple competitions we 
are able to determine if the 2022 data differed significantly from other 
competitions.
Our investigation into reaction times is important not only because if its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times; which is the goal of
every sprinter who steps onto the track.
%\eds{I recommend moving the rest of this paragraph to Section 2.2 as I don't
%think it is critical for readers to know all the details here.}
%First we can compare the reaction times of athletes who
%competed at the World Track and Field Championship in both 2019 and 2022, and 
%second we can compare how athletes performed at the 2022 championship relative 
%to their national/qualifiers meet. In order to reduce any chances
%of athletes improving reaction time, we can also look at data from qualifying meets.
%Almost every athlete had to qualify by proving they are elite at the national level, 
%usually by coming in at least fourth. Thus, data was compiled on athletes who competed
%at the 2022 World Track and Field Championships on what these athletes also ran 
%at their country's national championship. If there was any argument that athletes
%have improved their reaction time from 2019 to 2022, it should now be considered
%nearly null, as the national competitions took place between June and July 2022 
%and the World Championships were in August. The equipment used to measure reaction
%time cannot be guaranteed to be the same from national to international
%competition, but the time span between meets has been reduced considerably.

The rest of this paper is organized as follows. Section~\ref{sec:Data} describes 
how data was collected and begins to detail the generalized linear mixed-effect 
(GLMM) model that is further developed in Section~\ref{sec:Methods}. The 
signed-rank test that analyzes reaction times for the same athletes across
different competitions is
also described in Section~\ref{sec:Methods}.  Section~\ref{sec:Results}
 provides the results of the two analyses and ends 
by questioning the reaction time barrier.  Lastly, Section~\ref{sec:Discussion}
discusses the impact and limitations of the paper.
\eds{capitalize first letter when referring to a numbered section, table, figure, 
etc. ``Section 4 provides...'' vs. ``The next section provides...''}
\of{I think capitalization should be all set: all are in form of ``Section i''}

@@

Data was taken from the World Athletics 
%\eds{is there a website you can cite here?}
and covers the men's 110 meter hurdles from 1999 to 2022; and can be found in
the Appendix~\ref{sec:Appendix}.   \eds{see earlier comment - 
are you using any of these as covariates?}
\of{removed covariate confusion, talk about womens a couple sentences down} 
We chose not to include women's data in this
paper due to the numerous papers citing difference in reaction times in men and
women's reaction times: \citep{lipps2011implications, babicc2009reaction,
panoutsakopoulos2020gender}
The variable ``Stage" refers to whether the observation occurred during a 
%heat (preliminary round)  
semifinal heat (of which there are either two
or three each year) denoted by ``S", or final heat denoted by ``F".
The preliminary heats were not studied because athletes react faster as the 
competition progresses \citep[e.g.,][]{zhang2021correlation}. The variable 
``Gender" has levels ``M" and ``F" for ``male" and ``female" respectively.  The
variable ``Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly). There is no statistical meaning to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary heat of 2022 and Batch
111 corresponds to the 1999 finals). %Included are
%box plots of the reaction times of the men's finals and the men's semifinals and
%finals pooled together.  
For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together and refer to this data as the ``pooled data"
%We will refer to the latter as the ``pooled data"
throughout the paper.  %The heats or preliminary rounds were not studied because
 % Athletes react 
%faster in the finals than the semifinals which are faster than the heats 
%\citep{zhang2021correlation}.  
\eds{if you are really not using ``H'' data at all, it does not appear necessary
to spend time defining earlier.  You could just say directly when you define 
stage that you only consider semifinals and finals and cite the Zhang paper for 
rationale}
\of{Ok I updated this.  Should I delete the batch variable too?  Although I don't
use it, I do reference it in the methods section}
The pooled data increases the sample size which is advantageous because in some
years there are very few finals observations.  For example, in 2022 there
were only five data points due to two disqualifications and one athlete who 
did not compete.
The data under consideration is best visualized using side-by-side box plots, as
shown in Figure~\ref{fig:Boxplots}.  It is clear that the 2022 box plot for both 
the men's finals and semifinals data is lower (indicating faster reaction times)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
\citet{brosnan2017effects} found median reaction times of 0.156 seconds \eds{mean?  median?} 
\of{Resolved}
for men when looking at data from 1999 to 2014.

\eds{I wonder if this paragraph would be better placed in the introduction?}
\of{I think it could fit in the brief section between the start of the Data section
and the GLMM subsection in the paragraph starting with "There are two types..."}
Something that is extremely important to note in the data is the inconsistency 
of World Athletics in terms of how they penalize false starts.  From 2007--2009,
the IAAF (the former name for World Athletics, and the governing body for the 
World Track and Field Champions) instituted a rule change that allowed one 
warning false start before a sprinter was disqualified \citep{iaaf2009falsestart}. 
This was desirable for World Athletics as false starts can make already long 
track meets more tedious, both for athletes and viewers watching the television 
broadcast.  There were 18 male and 7 female false starts at the 2007 World 
Championships, and 18 male and 7 female starts at the 2009 World Championships. 
Starting in 2011, the rule was scrapped, and the old policy which automatically 
disqualifies runners who false start was reinstated. By returning to the harsher 
policy and cracking down on false starts, World Athletics had reduced men's 
false starts by two thirds \eds{for men, not overall} \of{Better?} in 2011 (6 male and 4 female disqualifications) 
\citep{iaaf2009falsestart}.
%The effect? At the 2011 
%World Championship there were 6 male disqualifications and 4 female 
%disqualifications \citep{iaaf2009falsestart}.  
%One study that looked at different false start rules that IAAF has
%imposed from 1997 to 2009 and found that the reaction times when there was less 
%leniency improved by a statistically significant amount
%\citep{haugen2013effect}. 
\citet{haugen2013effect} examined the effect of different false start
rules that IAAF has imposed from 1997 to 2009 and found statistically 
significant improvement in reaction times under more lenient rules.
\eds{is that right?  it seems counterintuitive to me}
\of{Yes you are correct, I updated the language.}

\eds{I am not sure the previous sentence follows from the one preceding it. Why
 does World Athletics find faster reaction times desirable?  Do you mean that
they find fewer false starts desirable?}

\eds{Two comments about the previous sentence: 1. 
it does not logically follow the previous sentence as there is no mention of 
gender, and 2. if you are not using women's data, don't spend time talking about 
it at the beginning of the data section.  Similar to my comment about 
preliminary heats, you can move this sentence to the beginning of the Data 
section to provide a rationale for why you do not include data for women.} 
\of{Updates: I moved any talk about women's data to start of Data section, I think
it flows much better now}

\subsection{Beyond World Championship}\label{sec:databeyond}
The motivation for looking beyond World Championship data came from an 
exploratory \eds{exploratory instead of qualitative?} \of{I agree; fixed}
analysis examining how United States athletes reacted at the 
2022 United States Track and Field Championships.
\eds{be consistent with how you refer to the US championship:
``United States Track and Field Championships" and ``United States Track and Field 
Championships" are both being used; please pick one and use that throughout}
\of{Resolved; changed all USA to United States.}
%This section was motivated from a qualitative analysis that started by looking
%at how United States athletes reacted at the 2022 United States Track and Field
%Championships. 
From June 23-26 2022, the United States held its Track
and Field Championship at Hayward 
Field in Eugene, Oregon to decide who to send to the World Championships 
being held at the same venue in August. %the exact same venue as used 
%for the World Track and Field Championship a month later.  
%Thus, a baseline is able to be established for the four United States athletes who 
Thus, we can establish a baseline for the four United States athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championship races compared to the United States Track and Field 
Championship races. Devon Allen raced three times at the United States Track and
Field Championship and three times at the World Track and Field Championship. 
His reaction times were substantially faster in the World Track and
Field Championship compared to the United States Track and Field Championship as were his
teammates. \eds{statistically significantly faster? If you used a statistical 
test, it should not be described in this section.  If you didn't use a 
statistical test, you should replace the word `significantly' with something 
else (substantially?)} \of{Replaced signficantly with substantially}
This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  Table~\ref{fig:USAvsWorld} highlights the difference in reaction 
times for Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. 
%``United States" denotes
%the United States Track and Field Championships, ``World" denotes the World Track and Field
%Championships. All numbers listed in the table are the reaction
%time for the respective athletes in seconds.

\begin{table}
\begin{center}
  \caption{Reaction times (in seconds) for four United States Athletes at the 
	2022 United States Track
  and Field World Championships (``United States") and the 2022 World Track and Field 
	Championships (``World"); H denotes preliminary heat, S denotes semifinal, and 
	F denotes final.
  ``N/A" was used for any athlete that did not compete and thus did not register 
  a reaction time. }
  \begin{tabular}{c c c c c c c} 
   \toprule
	 & \multicolumn{3}{c}{United States} & \multicolumn{3}{c}{World} \\
	\cmidrule(lr){2-4}
    \cmidrule(lr){5-7}
   Athlete &  H &  S &  F &  H &  S &  F \\ [0.5ex] 
   \midrule
   Devon Allen & 0.201 & 0.153 & 0.160 & 0.123 & 0.101 & 0.099 \\ 
   Trey Cunningham & 0.186 & 0.185 & 0.182 & 0.115 & 0.120 & 0.109 \\
   Grant Holloway & 0.192 & 0.190 & N/A & 0.147 & 0.128 & 0.124 \\
   Daniel Roberts & 0.181 & 0.200 & 0.183 & 0.179 & N/A & N/A \\ [0.5ex]
   \bottomrule
  \end{tabular}
  \label{fig:USAvsWorld}

@@

\of{Im not sure how to fix this but the spacing between the lines of the
table is different than the rest of the paper}

@@

To better understand this model, we can identify the gamma model in terms of the
commonly used shape and scale parameters. For a gamma distribution with
shape~$a$ and scale~$b$, we have mean $\mu = ab$ and variance $v = ab^2$. The
variance as a function of the mean $\mu$ in the GLM sense is
$v(\mu) = \mu^2 / a$, which implies that the dispersion parameter is
$\phi = 1 / a$. Therefore, for the gamma mixed-effect model, the conditional
gamma distribution of the reaction time $Y_{ij}$ given the venue effect $z_i$
has shape $a = 1 / \phi$ and scale $b = g^{-1}(\alpha + z_i) \phi$. The marginal
distribution of $Y_{ij}$ is a scale-mixture of gamma distributions, which can be
easily simulated from once the parameters are estimated. Many
random numbers generated from the fitted mixture distribution can be used to
approximate the probability of observing a reaction time faster than any given
threshold.  We are specifically interested in the probability of a reaction time
 being less than 0.1 seconds in order to gauge if that is a reasonable 
disqualification barrier.

As for the random effect, it was possible to include a heat 
\eds{heat? when I first read this, I thought you meant race as in ethnicity}
\of{Yes I think heat or batch effect works better, especially now that I have
removed most talk about heat from the data section} 
effect, as well.
From our exploration, however, the race effect is very difficult to explain or 
interpret as it means that some heats of races were either
faster or slower and that everyone in the race reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier, however, to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art. Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the race
effect, but are all factors that could potentially impact the venue effect.

@@

%Citation needed in this section
Since 1985 Seiko Holding Corporations has served every World Athletics 
Championship as the official timer \citep{wa2022seiko}.  Since 1985, the 
technology and the ability
to accurately predict measurements: long jump distances, false starts, total 
time, reaction time, etc. have all dramatically improved.  Seiko did not start 
tracking reaction time as an official measurement until 1999 in the men's and 
women's 100 meter \eds{110 for men?} hurdles. 
Seiko regularly updates their equipment so that they provide cutting edge
technology to the World Track and Field Championships, the highest stakes in the 
world of running outside of the Olympics. Seiko's technology for detecting 
reaction times relies on the pressure that athletes exert on the plate when they 
push off.  Their systems measure the time differential
between when the ``gun" goes off to start the race and when the pressure changes 
\citep{wa2022seiko}.  If an athlete has a reaction time under 0.1 seconds, it 
is deemed a 
false start as it is considered that no athlete can react so quickly 
\citep{Seiko-Timing}.  Thus, a reaction time of for example 0.05, suggests that 
the athlete predicted the gun and it was luck that caused their abnormally high 
time.  At the World Championship level, World Athletics and Seiko want to remove 
that element of luck and thus impose the 0.1 second barrier. In 2013, Seiko 
upgraded their timing equipment for sprinting events (110 meter hurdles falls 
under this category) \citep{wa2013backtage} \eds{110 for men?} \of{Fixed}. Since 1999, the
 highest \eds{highest as in fastest or slowest? Neither seems consistent with 
the boxplots in Figure 1} reaction 
times in the men's 110-meter hurdle were recorded in 2013.  That is not to say
 that the higher 
times were caused by Seiko's equipment, but the two may be related.  It is worth 
noting that prior
to 2022, Seiko again upgraded its technology but for its jump management system.
\eds{I would argue that this paragraph should go in the discussion as there are 
no results presented here}
